---
title: "Process and Add Assay Results to Database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Process and Add Assay Results to Database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Objective

This tutorial describes how to:

1. connect to the genetics run identification database with credentials
2. write plate run metadata to the database
3. prepare Synergy H1 result data for upload to the database
4. upload assay result data to the database
5. generate threshold values from the plate run
6. update assay detections for a plate run in the database

## Dependencies

The following packages are required:
```{r setup, message=FALSE, warning=FALSE}
library(grunID)
library(config)
library(DBI)
library(RPostgres)
```

## Database Credentials

For improved security, we recommend managing your database credential information 
with environment variables. Information on how to set up your **config.yml** file and
pull information from that file to establish a connection to the database can be found 
in the article `vignette("azure-authentication")` under **Authentication with a Config file**.

## Create Database Connection

To create a connection object, we run `gr_db_connect`:
```r
con <- gr_db_connect() # a config file will be searched for starting at the working directory.

dplyr::tbl(con, "agency")
```

You can also call `gr_db_connect` with arguments specifying your username, database name and host for the server for the database. Note that this is a quick way to create a connection to the database but if using source control its also an easy way to accidentally share your credentials. For more information on this, and to see code, please see the article `vignette("azure-authentication")` under **Authentication with username, dbname and host**.

Every time we use a function to read or write to the database, we will pass 
this connection object to the function.

## Add Plate Run Metadata

When a plate is run on the Synergy H1 reader, we want to relate critical plate run 
metadata with the assay results from that particular run in our database. We call 
`add_plate_run` providing the protocol ID, method ID, laboratory ID, and the 
person who ran the plate. The function returns a unique plate run ID that will be
associated with the assay result records. Because our assay results reference the
plate run ID, we must add a plate run in our database before adding assay result
data.

If you do not know the ID for the protocol used, you can review all existing 
protocols stored in the database with the following code: 
```{r, eval=FALSE, cache=TRUE}
all_protocols <- grunID::get_protocol(con)
```

Select appropriate protocol ID with this code:
```{r, eval=FALSE, cache=TRUE}
protocol_id <- all_protocols |>
  filter(id == 1) |>
  pull(id)
```
Similarly, you can call `get_genetic_method` and `get_laboratory_id` to
review the IDs associated with methods and laboratories within the database.

```{r, eval=FALSE, cache=TRUE}
genetic_method_id <- get_genetic_methods(con) |>
  filter(method_name == "SHERLOCK") |>
  pull(id) # SHERLOCK

laboratory_id <- get_laboratories(con) |>
  filter(stringr::str_detect(code, "DWR")) |> pull(id) # DWR GeM

# store output of add_plate_run - contains unique plate run ID needed for later functions
plate_run_id <- grunID::add_plate_run(con, protocol_id, genetic_method_id,
                                      laboratory_id, lab_work_performed_by = "Yogi Bear",
                                      description = "Plate run for sampling event X",
                                      date_run = "07-04-1776")
```

**Note: save the output of `add_plate_run_id`. This contains the unique plate run identifier and will be needed
to pass to later functions**

## Prepare Sample Details

The Synergy H1 result output from plate runs is organized using generic identifiers 
that need to be mapped to the real sample identifiers in order to properly associate
the results with the correct sample id in the database. The output file’s **Layout** 
section contains the well locations for samples encoded as non-unique generic 
sample ids *(e.g., SPL1, BLK)*. The number of rows and columns in the **Layout** vary based on the plate size. A plate size 96 has rows A:H and columns 1:12 and a 
plate size 384 has rows A:P and columns 1:24. The results for each sample are found 
within columns that are the letter-number combination of the well location captured 
in the **Layout** section *(e.g., if SPL1 is in row A and column 3, then the measurements at each time step are recorded in a column labeled A3)*. To map the sample ids to 
the generic identifiers produced by the Synergy H1 software *(e.g., SPL1 and A3)*, 
we must provide the true sample ids and other metadata at each well location
in the results as a dataframe containing these columns:

* **location:** *(character)* letter-number combination based on Synergy H1 Layout grid (e.g., A3)
* **sample_id:** *(character)* unique ID of mucus swab or fin clip made up of site code, year, sample event, sampling bin, and count within sampling bin (e.g., BTC22_3_A_1)
* **sample_type:** *(numeric)* provide 1 for mucus and 2 for fin clip
* **assay_id:** *(numeric)* the ID for assay types (e.g., 1 = OTS28 Early 1), use `get_assay` to view all available options.
* **plate_run_id:** *(numeric)* the ID returned from running `add_plate_run` to 
associate important metadata with the assay results

We can map the generic sample identifiers *(e.g. SPL1)* to the JPE sample identifiers
*(e.g. F1722_3_A_1)* using the `plate_map` tab of the output file. This tab should
contain the well layout locations with JPE sample identifiers, and the name of the 
tab has to be `plate_map`. The function `process_well_sample_details` then reads 
in that layout and maps the JPE sample identifiers to well locations, producing
a dataframe in the correct format:

```{r, eval = FALSE, cache = TRUE}
sample_details <- process_well_sample_details(filepath = "templates/sherlock_results_template.xlsx",
                                              sample_type = "mucus",
                                              layout_type = "split_plate_early_late",
                                              plate_run_id = plate_run_id)
```

The object sample_details is now a data frame that looks like:
```{eval=FALSE}
location,sample_id,sample_type_id,assay_id,plate_run_id
A3,BTC22_3_A_1,1,1,1
A4,BTC22_3_A_2,1,1,1
A6,BTC22_3_A_3,1,1,1
B3,BTC22_3_B_1,1,1,1
B4,BTC22_3_B_2,1,1,1
B6,BTC22_3_B_3,1,1,1
```

**The sample details as a dataframe are an argument to `process_sherlock`, the function
that extracts assay result data from the Synergy H1 reader output and transforms
the data into the correct format for the database.** This maps the generic sample 
identifiers to the JPE sample identifiers by well location.

## Process SHERLOCK Results

To extract and transpose the assay result data from the Synergy H1 reader output,
provide the `process_sherlock` function the filepath to the result data excel workbook,
the sample details dataframe described above, and the plate size run *(either 96 or 384)*.
The function will return a list containing both the raw and final assay results.
```{r, eval=FALSE, cache=TRUE}
results <- grunID::process_sherlock(sherlock_results_filepath = "templates/sherlock_results_template.xlsx",
                                    sample_details = sample_details,
                                    plate_size = 384)
```

**Note: save the output of `process_sherlock`. This contains the data that will be needed to
to pass to later functions.**

## Upload Assay Results

Provide `add_assay_results` a valid connection to the database and the processed
assay result data. The function returns the number or records added.
```{r, eval=FALSE, cache=TRUE}
grunID::add_assay_results(con, results)
```

**Note: you do not need to save the output of `add_assay_results`.**

## Generate Threshold Values

Provide `generate_threshold` a valid connection to the database and the plate run id. The function calculates
the raw fluorescence threshold values for each assay on a plate run. Currently, this is calculated as two times the mean value
of the last time step from the control blank wells.
The function returns a table containing thresholds for a sampling event and needs to be passed to `update_assay_detections` to assign
detections for assays.

```{r, eval=FALSE, cache=TRUE}
thresholds <- grunID::generate_threshold(con, plate_run_id)
```
**Note: save the output of `generate_threshold`. This contains the threshold values and needs to be passed to `update_assay_detection`**

## Update Assay Detections
Provide `update_assay_detection` a valid connection to the database and the thresholds calculated for a plate run using `generate_threshold`. The function updates the assay result table with positive detections and, if appicable, the genetic run type identification.

```{r, eval=FALSE, cache=TRUE}
grunID::update_assay_detection(con, thresholds_event)
```
**Note: you do not need to save the output of `update_assay_detection`.**

## End Session

And finally, close your open connection to the database:
```{r, eval=FALSE, cache=TRUE}
DBI::dbDisconnect(con)
```

## Summary

Here is the code again to:

1. securely manage database credentials
2. connect to the genetics run identification database
3. write plate run metadata to the database
4. prepare Synergy H1 result data for upload to the database
5. upload assay result data to the database
6. generate threshold values from the plate run
7. update assay detections for a plate run in the database

```{r, eval=FALSE, cache=TRUE}
con <- gr_db_connect()

plate_run_id <- grunID::add_plate_run(con, protocol_id, genetic_method_id,
                                      laboratory_id, lab_work_performed_by = "Yogi Bear",
                                      description = "Plate run for sampling event X",
                                      date_run = "07-04-1776")

sample_details <- process_well_sample_details(filepath = "templates/sherlock_results_template.xlsx",
                                              sample_type = "mucus",
                                              layout_type = "split_plate_early_late",
                                              plate_run_id = plate_run_id)

results <- grunID::process_sherlock(sherlock_results_filepath = "templates/sherlock_results_template.xlsx",
                                    sample_details = sample_details,
                                    plate_size = 384)

grunID::add_assay_results(con, results)

thresholds <- grunID::generate_threshold(con, plate_run_id)

grunID::update_assay_detection(con, thresholds_event)

DBI::dbDisconnect(con)
```
