---
title: "Process and Add Assay Results to Database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Process and Add Assay Results to Database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Objective

This tutorial describes how to:

1. connect to the genetics run identification database with credentials
2. write plate run metadata to the database
3. prepare Synergy H1 result data for upload to the database
4. upload assay result data to the database
5. generate threshold values from the plate run
6. update assay detections for a plate run in the database

## Dependencies

The following packages are required:
```{r setup, message=FALSE, warning=FALSE}
library(grunID)
library(config)
library(DBI)
library(RPostgres)
```

## Database Credentials

For improved security, we recommend managing your database credential information 
with environment variables. Information on how to set up your **config.yml** file and
pull information from that file to establish a connection to the database can be found 
in the article `vignette("azure-authentication")` under **Authentication with a Config file**.

## Create Database Connection

To create a connection object, we run `gr_db_connect`:
```r
con <- gr_db_connect() # a config file will be searched for starting at the working directory.

dplyr::tbl(con, "agency")
```

You can also call `gr_db_connect` with arguments specifying your username, database name and host for the server for the database. Note that this is a quick way to create a connection to the database but if using source control its also an easy way to accidentally share your credentials. For more information on this, and to see code, please see the article `vignette("azure-authentication")` under **Authentication with username, dbname and host**.

Every time we use a function to read or write to the database, we will pass 
this connection object to the function.

## Add Plate Run Metadata

When a plate is run on the Synergy H1 reader, we want to relate critical plate run 
metadata with the assay results from that particular run in our database. We call 
`add_plate_run` providing the protocol ID, method ID, laboratory ID, and the 
person who ran the plate. The function returns a unique plate run ID that will be
associated with the assay result records. Because our assay results reference the
plate run ID, we must add a plate run in our database before adding assay result
data.

If you do not know the ID for the protocol used, you can review all existing 
protocols stored in the database with the following code: 
```{r, eval=FALSE, cache=TRUE}
all_protocols <- grunID::get_protocols(con)
```

Select appropriate protocol ID with this code:
```{r, eval=FALSE, cache=TRUE}
protocol_id <- all_protocols |>
  filter(id == 1) |>
  pull(id)
```
Similarly, you can call `get_genetic_method` and `get_laboratory_id` to
review the IDs associated with methods and laboratories within the database.

```{r, eval=FALSE, cache=TRUE}
genetic_method_id <- get_genetic_methods(con) |>
  filter(method_name == "SHERLOCK") |>
  pull(id) # SHERLOCK

laboratory_id <- get_laboratories(con) |>
  filter(stringr::str_detect(code, "DWR")) |> pull(id) # DWR GeM

# store output of add_plate_run - contains unique plate run ID needed for later functions
plate_run <- grunID::add_plate_run(con, protocol_id, genetic_method_id,
                                      laboratory_id, lab_work_performed_by = "Yogi Bear",
                                      description = "Plate run for sampling event X",
                                      date_run = "07-04-1776")
```


**Note: save the output of `add_plate_run`. This contains the unique plate run identifier and will be needed
to pass to later functions**

## Prepare Sample Details and Sherlock Output

The SHERLOCK resusts file must be formatted into the appropriate excel sheet format, you can view an example
of this [here](https://srjpe.github.io/grunID/articles/grunID-data-templates.html). To add results to the
database you must have a "plate_run" associated with this upload. In this example we will use the plate run
we created in the previous step. You can query for existing plate run using the `get_plate_run` function.

## Process SHERLOCK Results

To extract and transpose the assay result data from the Synergy H1 reader output,
provide the `process_sherlock` function the filepath to the result data excel workbook,
the sample details dataframe described above, and the plate size run *(either 96 or 384)*.
The function will return a list containing both the raw and final assay results.

```{r, eval=FALSE, cache=TRUE}
results <- grunID::process_sherlock(filepath = "templates/sherlock_results_template.xlsx",
                                    sample_type = "mucus",
                                    layout_type = "split_plate_early_late",
                                    plate_run_id = plate_run
                                    plate_size = 384)
```

**Note: save the output of `process_sherlock`. This contains data that will be needed to
to pass to `add_raw_assay_results`.**

## Upload Assay Results

Provide `add_raw_assay_results` a valid connection to the database and the processed
assay result data. The function returns the number or records added.
```{r, eval=FALSE, cache=TRUE}
grunID::add_raw_assay_results(con, results)
```

**Note: you do not need to save the output of `add_raw_assay_results`.**

## Generate Threshold Values

Provide `generate_threshold` a valid connection to the database and the plate run id. The function calculates
the raw fluorescence threshold values for each assay on a plate run. Currently, this is calculated as two times the mean value
of the last time step from the control blank wells.
The function returns a table containing thresholds for a sampling event and needs to be passed to `update_assay_detections` to assign
detections for assays.

```{r, eval=FALSE, cache=TRUE}
thresholds <- grunID::generate_threshold(con, plate_run_id)
```
**Note: save the output of `generate_threshold`. This contains the threshold values and needs to be passed to `update_assay_detection`**

## Update Assay Detections
Provide `update_assay_detection` a valid connection to the database and the thresholds calculated for a plate run using `generate_threshold`. The function updates the assay result table with positive detections and, if appicable, the genetic run type identification.

```{r, eval=FALSE, cache=TRUE}
grunID::update_assay_detection(con, thresholds_event)
```
**Note: you do not need to save the output of `update_assay_detection`.**


## End Session

And finally, close your open connection to the database:
```{r, eval=FALSE, cache=TRUE}
DBI::dbDisconnect(con)
```

## Summary

Here is the code again to:

1. connect to the genetics run identification database with credentials
2. write plate run metadata to the database
3. prepare Synergy H1 result data for upload to the database
4. upload assay result data to the database
5. generate threshold values from the plate run
6. update assay detections for a plate run in the database

```{r, eval=FALSE, cache=TRUE}
con <- gr_db_connect()

plate_run_id <- grunID::add_plate_run(con, protocol_id, genetic_method_id,
                                      laboratory_id, lab_work_performed_by = "Yogi Bear",
                                      description = "Plate run for sampling event X",
                                      date_run = "07-04-1776")

results <- grunID::process_sherlock(filepath = "templates/sherlock_results_template.xlsx",
                                    sample_type = "mucus",
                                    layout_type = "split_plate_early_late",
                                    plate_run_id = plate_run_id
                                    plate_size = 384)

# no need to store output
grunID::add_raw_assay_results(con, results)

thresholds <- grunID::generate_threshold(con, plate_run_id)

# no need to store output
grunID::update_assay_detection(con, thresholds)

# disconnect
DBI::dbDisconnect(con)
```
