---
title: "Process and Add Assay Results to Database"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Process and Add Results to Database}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Objective

This tutorial describes how to:

1. securely manage database credentials
2. connect to the genetics run identification database
3. write plate run metadata to the database
4. prepare Synergy H1 result data for upload to the database
5. upload assay result data to the database

## Dependencies

The following packages are required:
```{r setup, message=FALSE, warning=FALSE}
library(grunID)
library(config)
library(DBI)
library(RPostgres)
```

## Database Credentials

For improved security, we recommend managing your database credential information 
with environment variables. The `config` package makes it easy to manage environment 
variables. The database administrator will provide you with the necessary credential 
information to connect to the database.

First, you will need to create a **config.yml** file with the following information:

```{yaml, eval=FALSE}
default:
  dbname: {DATABASE_NAME}
  host: {HOST_ADDRESS}
  port: {PORT_NUMBER}
  username: {YOUR_USERNAME}
  password: {YOUR_PASSWORD}
```
Replace all `{*}` place holders with the credential information provided by the 
database administrator.

It is important to not accidentally share credentials online. If you are pushing 
your analysis to Github make sure to add `config.yml` to your **.gitignore** file.

By default, the configuration data is read from **config.yml** within the current
working directory (or parent directories if no config file is found in the initially 
specified directory). For more details see the [config docs](https://rstudio.github.io/config/).

To load the credential information into memory call:
```{r, eval=FALSE}
cfg <- config::get()
```

## Database Connection

The `DBI` package provides simple and consistent methods for connecting R to 
database management systems (DBMS). For more details see the [DBI docs](https://dbi.r-dbi.org/). 
The genetic run identification DBMS is implemented with PostgreSQL. We use the 
`RPostgres` package to define the proper operations for creating connections and 
defining data type mappings. For more details see the [RPostgres docs](https://rpostgres.r-dbi.org/).

To create a connection object, we pass our credentials to `DBI::dbConnect()` and 
specify a PostgreSQL driver.

```{r, eval=FALSE, cache=TRUE}
con <- DBI::dbConnect(RPostgres::Postgres(),
                      dbname = cfg$dbname,
                      host = cfg$host, 
                      port = cfg$port,
                      user = cfg$username,
                      password = cfg$password)
```
Every time we use a function to read or write to the database, we will pass 
this connection object to the function.

## Add Plate Run Metadata

When a plate is run on the Synergy H1 reader, we want to relate critical plate run 
metadata with the assay results from that particular run in our database. We call 
`add_plate_run()` providing the protocol ID, method ID, laboratory ID, and the 
person who ran the plate. The function returns a unique plate run ID, that we will 
include with the assay result records. Because our assay results reference the
plate run ID, we must add a plate run in our database before adding assay result
data.

If you do not know the ID for the protocol used, you can review all existing 
protocols stored in the database with the following code: 
```{r, eval=FALSE, cache=TRUE}
all_protocols <- grunID::get_protocol(con)
```

Select appropriate protocol ID with this code:
```{r, eval=FALSE, cache=TRUE}
protocol_id <- all_protocols[1, "id", drop = TRUE]
```
Similarly, you can call `get_genetic_method()` and `get_laboratory_id()` to
review the IDs associated with methods and laboratories within the database.

```{r, eval=FALSE, cache=TRUE}
genetic_method_id <- 1 # SHERLOCK
laboratory_id <- 1 # DWR GeM
plate_run_uid <- grunID::add_plate_run(con, protocol_id, genetic_method_id,
                                       laboratory_id, lab_work_preformed_by = "Yogi Bear")
```

## Prepare Sample Details

The Synergy H1 result output from plate runs is organized using generic identifiers 
that need to be mapped to the real sample identifiers in order to properly associate
the results with the correct sample id in the database.The output file’s **Layout** 
section contains the well locations for samples encoded as non-unique generic 
sample ids *(e.g., SPL1, BLK)*. The number of rows and columns in the **Layout** vary 
based on the plate size. A plate size 96 has rows A:H and columns 1:12 and a 
plate size 384 has rows A:P and columns 1:24. The results for each sample are found 
within columns that are the letter-number combination of the well location captured 
in the **Layout** section *(e.g., if SPL1 is in row A and column 3, then the measurements at each time step are recorded in a column labeled A3)*. To map the sample ids to 
the generic identifiers produced by the Synergy H1 software *(e.g., SPL1 and A3)*, 
we must provide the true sample ids and other metadata at each well location
in the as a dataframe containing these columns:

* **location:** *(character)* letter-number combination based on Synergy H1 Layout grid (e.g., A3)
* **sample_id:** *(character)* unique ID of mucus swab or fin clip made up of site code, year, sample event, sampling bin, and count within sampling bin (e.g., BTC22_3_A_1)
* **sample_type:** *(numeric)* provide 1 for mucus and 2 for fin clip
* **assay_id:** *(numeric)* the ID for assay types (e.g., 1 = OTS28 Early 1), use `get_assay()` to view all available options.
* **plate_run_id:** *(numeric)* the ID returned from running `add_plate_run()`, to 
associate important metadata with the assay results

Here is an **example.csv** file that is in the correct format for capturing sample details.
```{eval=FALSE}
location,sample_id,sample_type_id,assay_id,plate_run_id
A3,BTC22_3_A_1,1,1,1
A4,BTC22_3_A_2,1,1,1
A6,BTC22_3_A_3,1,1,1
B3,BTC22_3_B_1,1,1,1
B4,BTC22_3_B_2,1,1,1
B6,BTC22_3_B_3,1,1,1
```

The sample details as a dataframe are an argument to `process_sherlock()`, the function
that extracts assay result data from the Synergy H1 reader output and transforms
the data into the correct format for the database.
```{r, eval=FALSE, cache=TRUE}
sample_details <- readr::read_csv("path_to/example.csv")
```

## Process SHERLOCK Results

To extract and transpose the assay result data from the Synergy H1 reader output,
provide the `process_sherlock()` function the filepath to the result data excel workbook,
the sample details dataframe described above, and the plate size run *(either 96 or 384)*.
The function will return a list containing both the raw and final assay results.
```{r, eval=FALSE, cache=TRUE}
results <- grunID::process_sherlock(sherlock_results_filepath = "path_to/SynergyH1_data.xlsx",
                                    sample_details = sample_details,
                                    plate_size = 96)
```

## Upload Assay Results

Provide `add_assay_results` a valid connection to the database and the processed
assay result data. The function returns the number or records added.
```{r, eval=FALSE, cache=TRUE}
grunID::add_assay_results(con, results)
```

And finally, close your open connection to the database:
```{r, eval=FALSE, cache=TRUE}
DBI::dbDisconnect(con)
```

## Summary

Here is the code again to:

1. securely manage database credentials
2. connect to the genetics run identification database
3. write plate run metadata to the database
4. prepare Synergy H1 result data for upload to the database
5. upload assay result data to the database

```{r, eval=FALSE, cache=TRUE}
cfg <- config::get()

con <- DBI::dbConnect(RPostgres::Postgres(),
                      dbname = cfg$dbname,
                      host = cfg$host,
                      port = cfg$port,
                      user = cfg$username,
                      password = cfg$password)

plate_run_uid <- grunID::add_plate_run(con, protocol_id = 1, genetic_method_id = 1,
                                       laboratory_id = 1, lab_work_preformed_by = "Yogi Bear")

sample_details <- readr::read_csv("path_to/example.csv")

results <- grunID::process_sherlock(sherlock_results_filepath = "path_to/SynergyH1_data.xlsx",
                                    sample_details = sample_details,
                                    plate_size = 96)

grunID::add_assay_results(con, results)

DBI::dbDisconnect(con)
```
