---
title: "Querying Sample Status"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Querying Sample Status}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The following packages are required:

```{r, message=FALSE, warning=FALSE}
library(grunID)
library(tidyverse)
```

## Objective

This tutorial describes how to:

1) Query the database for the status of a specific sample ID (or multiple sample IDs)
2) Query the database for all sample IDs requiring further analysis

## Check Status of Specific Samples

The function `get_sample_status` is helpful for when you already know the sample IDs about which you want more information. The function takes in one or more sample IDs and a connection and returns the following information:

* id (unique identifier generated for the sample status table)
* sample_id
* status_code_id
* status_code_name
* comment
* created_at
* created_by

The function defaults to `full_history = FALSE`, showing you only the current status of the sample. If you want the full history of a sample, you can change the argument to `full_history = TRUE`.

```{r, eval = FALSE, cache = TRUE}
con <- grunID::gr_db_connect()

# find the status of a specific sample "DER23_1_C_4" without the full history
grunID::get_sample_status(con, "DER23_1_C_4", full_history = FALSE)

# find the status of a specific sample "DER23_1_C_4" WITH the full history
grunID::get_sample_status(con, "DER23_1_C_4", full_history = TRUE)

```


## Get a List of All Samples Needing Analysis

If you don't have a specific sample ID in mind but want to know which samples in the database need further assays or clarification, you can use the function `get_samples_needing_action`. This function produces a list of Sample IDs organized into the following categories: 

* needs_ots_16 
* repeat_ots_28 
* repeat_ots_16
* potential_heterozygotes

```{r, eval = FALSE, cache = TRUE}
# produce a list of Sample IDs organized by the analysis required
grunID::get_samples_needing_action(con)
```


